\thetitle

### 2021-03-01 writing/rational-effort.tex
rational programmer with editorial voice

In addition to successful and failure information for the various trails,  
our experiment  records the number of components a rational programmer 
has to type along each trail ($\lvert \conf_n \setminus \conf_0
\rvert$). We use this number as the metric for the effort 
of the rational programmer to debug an interesting scenario.

A comparison of the effort of different modes of the rational programmer
can help shed some further light to the effectiveness of the
three gradual typing systems. For example, consider one debugging scenario.  
If, for instance, both  Natural blame and Transient first
blame trails are successful, the two modes of the 
rational programmer can compete to see which one
debugs the interesting scenario with less effort. In
general, if the effort distribution for a mode of the rational programmer
has a shorter tail and more
volume around smaller values compared to the effort distribution of another
mode, then this is evidence that the first mode is the more effective of the two.

Finally, effort allows us to test whether the observed effectiveness 
of the rational programmer is an artifact of pure chance or not.
That is, we can compare the effort distribution for a mode of the
rational programmer with another mode that ignores error information entirely
and instead selects which component to type next randomly. 
Of course, since each mutant has a finite number of components, 
such random mode trails are always successful. At the same time though,
the Natural, Transient, and Erasure random modes exhibit a normal effort
distribution across scenarios that should be distinct from the effort distribution for all other modes of
the rational programmer if the effectiveness of the modes are not
coincidental.

>>>>

%% -----------------------------------------------------------------------------

In addition to success and failure information for the various trails, the
experiment software records the number of components a rational programmer has
to equip with types along each trail ($\lvert \conf_n \setminus \conf_0
\rvert$). This number can serve as a metric.

A comparison of the effort of different modes of the rational programmer can
illuminate the comparison of the effectiveness of the three gradual typing
systems. For example, consider a debugging scenario.  If, for instance, both
Natural blame and Transient first blame trails are successful, the two modes of
the rational programmer can compete to see which one debugs the interesting
scenario with less effort. In general, if the effort distribution for a mode of
the rational programmer has a shorter tail and more volume around smaller values
compared to the effort distribution of another mode, then the first mode is
probably the more effective of the two.

Measuring effort can also tell whether the observed effectiveness of the
rational programmer is an artifact of pure chance or not.  Doing so means
comparing the effort distribution for a mode of the rational programmer with
another mode that ignores error information entirely and instead selects which
component to type next randomly.  Since each mutant has a finite number of
components, such random mode trails are naturally always successful. At the same
time though, the Natural, Transient, and Erasure random modes exhibit a normal
effort distribution across scenarios that should be distinct from the effort
distribution for all other modes of the rational programmer if the effectiveness
of the modes are not coincidental.

### 2021-03-01 writing/rational-experiment.tex
rational programmer with editorial voice

Trails and their properties give us the tools for a rigorous examination
of blame for Natural, Transient and Erasure Typed Racket in the setting of the 
mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
our experimental process collects data to answer three initial questions
for our interesting debugging scenarios:
\begin{itemize}
\item[$Q_1$] Is blame useful in the context of Natural?

@@

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example, experimental
question $Q_1$ asks whether blame is valuable for Natural and our experiment
uses the Natural blame and exception trails to answer it.

@@

In detail the answer to $Q_1$ demands a comparison of the success of the 
Natural blame and Natural exception trails for all interesting 
debugging scenarios.
  The first step is to construct each
mutant's scenario lattice and identify their interesting debugging
scenarios.
Our test
bed extends the trails that start from such roots according to the Natural-blame programmer.  If no scenarios
can be added to the trail, we check if
the last scenario of the trail type-checks or not. If it does, we record that the Natural-blame
trail is successful, and otherwise that it fails. We repeat the process for the
same roots again but for the other mode, Natural exceptions.
 Figure~\ref{fig:process}
summarizes this
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutate}.
After completing the experiment, we calculate the
success/failure results of the trails to determine for each root whether Natural blame
is more useful than Natural exceptions. We
obtain a positive answer for
$Q_1$ when a root exists where the above is true since it is evidence that
there is at least one interesting scenario that the rational programmer
manages to debug with blame information. 
The process is analogous for $Q_2$ and $Q_3$, using the respective
modes of the rational programmer.

For $Q_*$, the process is a bit more
involved. At a first level we proceed in a similar
manner except that we use two modes 
of the rational programmer from different semantics.
However, answering this question requires us to compare the percentage of scenarios
where one mode is more useful than the other and the inverse.
For instance, to
decide whether blame for Natural is more useful than first blame for Transient, 
we compare the percentage of interesting scenarios where 
Natural blame is more successful than Transient first blame 
with the percentage of interesting scenarios where Transient first blame 
is more successful than Natural blame. 
Finally we repeat these steps one more time to compare Natural blame and Transient last blame
and get a complete picture of the comparative usefulness of blame
in the two semantics.

>>>>

Trails and their properties provide the tools for a rigorous examination of
blame for Natural, Transient and Erasure. In line with the discussion so far,
the experiment software collects data to answer three initial questions for 
interesting debugging scenarios:
\begin{itemize}
\item[$Q_1$] Is blame useful in the context of Natural?

@@

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example, experimental
question $Q_1$ asks whether blame is valuable for Natural and the experiment
uses the Natural blame and exception trails to answer it.

@@

In detail the answer to $Q_1$ demands a comparison of the success of the Natural
blame and Natural exception trails for all interesting debugging scenarios.  The
first step is to construct each mutant's scenario lattice and identify their
interesting debugging scenarios.  The experiment software extends the trails that start
from such roots according to the Natural-blame programmer.  If no scenarios can
be added to the trail, the experiment software checks whether the last scenario
of the trail type-checks or not. If it does, the software records that the
Natural-blame trail is successful; otherwise it is recorded as failing. The
process is repeated for the same roots again but for the other mode, Natural
exceptions.  Figure~\ref{fig:process} summarizes this experimental process for
one mode of the rational programmer and connects it with the mutations from
section~\ref{sec:mutate}.  After completing the experiment, the experiment
software reports the success/failure results of the trails to determine for each
root whether Natural blame is more useful than Natural exceptions.  Question
$Q_1$ has a positive answer if a root exists where the above is true because it
is evidence that there is at least one interesting scenario that the rational
programmer manages to debug with blame information.  The process is analogous
for $Q_2$ and $Q_3$, using the respective modes of the rational programmer.

For $Q_*$, the process is a bit more involved. Answering this question calls for
a comparison of the percentage of scenarios where one mode is more useful than
the other and the inverse.  For instance, to decide whether blame for Natural is
more useful than first blame for Transient, the software compares the percentage
of interesting scenarios where Natural blame is more successful than Transient
first blame with the percentage of interesting scenarios where Transient first
blame is more successful than Natural blame.  Finally these steps are repeated
one more time to compare Natural blame and Transient-last blame and get a
complete picture of the comparative usefulness of blame in the two semantics.

### 2021-03-01 writing/rational-natural.tex
factor the rational programmer
rational programmer with editorial voice

%% -----------------------------------------------------------------------------
\subsection{Modes of the Natural Rational Programmer} \label{sub:natural}

@@

We can turn this description into a formal definition of the trail. 
\begin{quote}
\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
$\blame{\system, \conf}$ denotes the component that $\conf$ blames.
\end{quote}

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff its last scenario $\conf_n$ does not type check.  A Natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is, failing Natural blame trails are those that have not reached a scenario that
reveals the bug at compile time and their last scenario does not blame an untyped component. Thus
the rational programmer gets no hints from the gradual type system on how to
continue the search for the bug.

@@

To account for this situation, we devise a new mode of the Natural rational
programmer that follows a migration process based on exceptions:
\begin{quote}
\it A Natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$
where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
\end{quote}

>>>>

%% -----------------------------------------------------------------------------

@@

Here is a rigorous definition.
\begin{quote}
\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
$\blame{\system, \conf}$ denotes the component that $\conf$ blames.
\end{quote}

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff (the program for) its last scenario $\conf_n$ does not type check.  A Natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff (the program for) $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is, failing Natural blame trails are those that have not reached a scenario
that reveals the bug at compile time and ends in a scenario that does not blame
an untyped component during program execution. Thus the rational programmer gets
no hints from the gradual type system on how to continue the search for the bug.

@@

To account for this situation, we devise a new mode of the Natural rational
programmer that follows a migration process based on exceptions:
\begin{quote}
\it A {\em Natural exception trail\/} is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$
where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
\end{quote}

### 2021-03-01 writing/rational-stuff.tex
rational programmer with editorial voice

\citet{tfgnvf-popl-2016} describe the set of all possible type migrations with a
lattice. Formally, a program $\system$ is a set of modules $\set{\component}$.
Let a configuration $\conf$ of $\system$ be the subset of $\set{\component}$ that
comes with (possibly type-incorrect) annotations. These configurations of
$\system$ are ordered by the subset relation and form a lattice
$\lattice{\system}$ with $2^{\size{\system}}$ elements. The bottom scenario of
$\lattice{\system}$ is $\emptyset$, the top one $\system$ itself. In between are
the mixed-typed scenarios.

@@

The actions of the rational programmer create an ascending chain---dubbed a
\emph{trail}---in $\lattice{\system}$.  The starting scenario $\conf_0$ of a
trail---also referred to as the \emph{root} (of the trail)---is the initial
interesting debugging scenario.  Starting from such a root, the rational
programmer steps along a trail. If the scenario type-checks, the programmers runs
the program until it raises a run time error.  The rational programmer uses the
information in the error to decide which component(s) to equip with types. This
choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus creates a
trail. The trail ends when programmer reaches a scenarios that contain the mutant
module, because the type checker rejects it outright. At this point, the
programmer has identified the source of the impedance mismatch.

>>>>

\citet{tfgnvf-popl-2016} describe the set of all possible type migrations with a
lattice. Formally, a program $\system$ is a set of modules $\set{\component}$.
Let a configuration $\conf$ of $\system$ be the subset of $\set{\component}$
that comes with type annotations. These configurations of $\system$ are ordered
by the subset relation and form a lattice $\lattice{\system}$ with
$2^{\size{\system}}$ elements. The bottom scenario of $\lattice{\system}$ is
$\emptyset$, the top one $\system$ itself; in between are the mixed-typed
scenarios.

@@

The actions of the rational programmer create an ascending chain---dubbed a
\emph{trail}---in $\lattice{\system}$.  The starting scenario $\conf_0$ of a
trail---also referred to as the \emph{root} (of the trail)---is the initial
interesting debugging scenario.  Starting from such a root, the rational
programmer steps along a trail. If the (program for this) scenario type-checks,
the program is run until it raises a run-time error.  The rational programmer
uses the information in the error to decide which component to equip with
types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus
creates a trail. The trail's construction ends when it reaches a scenarios that
contain the mutant module because the type checker rejects its typed version
outright.  At this point, the source of the impedance mismatch is identified.

### 2021-03-01 writing/rational-transient.tex
rational programmer with editorial voice

The transient semantics assigns blame to a set of components. In this setting, a
blame assignment has the following meaning: the witness value to the impedance
mismatch may have crossed the boundaries between elements in the set, and each crossing
checked the value's type in a shallow manner.

This ambiguity in Transient blame raises the question of how the rational programmer
should react when the language produces a blame set. Our answer is that the rational
programmer has at least two reasonable options. The first one is to select the component that is added
to the blame set first and assign types to only that one---after all, if
fully checked, the types of
this first component should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones. The second option is to
select the component that is added to the blame set last, effectively
interpreting the blame set as a boundary-aware stack.

These two modes of rationalizing give rise to two different notions of trail.
\begin{quote}
\it A \emph{Transient-first blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
first component Transient adds to the blame set for $\conf$.
\end{quote}

and

\begin{quote}
\it A \emph{Transient-last blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
  \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
last component Transient adds to the blame set for $\conf$.
\end{quote}

The definition of \emph{Transient exception trails} 
is analogous to that for Natural and, same as for Natural, we use it to  
specify when  
Transient-first and Transient-last blame are more useful than 
Transient exceptions for debugging a scenario.

>>>>

The transient semantics assigns blame to a sequence of components. The blame
assignment says that the value witnessing the impedance mismatch may have
crossed the boundaries between elements in the sequence, and each crossing
checked the value's type in a shallow manner.

This ambiguity in Transient blame raises the question of how the rational
programmer should react when the language produces a blame sequence. Our answer
is that the rational programmer has at least two reasonable options. The first
one is to select the component that is added to the blame sequence first and
assign types to only that one---after all, if fully checked, the types of this
first component should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones. The second option is to select the
component that is added to the blame sequence last, effectively interpreting the
blame sequence as a boundary-aware stack.

These two modes of rationalizing give rise to two different notions of trail.

\begin{quote}
\it A \emph{Transient-first blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
first component Transient adds to the blame sequence for $\conf$.
\end{quote}

\begin{quote}
\it A \emph{Transient-last blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
  \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
last component Transient adds to the blame sequence for $\conf$.
\end{quote}

The definition of \emph{Transient exception trails} is analogous to that for
Natural. It is used to specify when Transient-first and Transient-last blame are
more useful than Transient exceptions for debugging a scenario.

### 2021-03-01 writing/rational.tex
stuff
name files properly
re-established the old organization
rational programmer with editorial voice

Section~\ref{sec:why-rational} explains, and illustrates with an example, how a
migratory type setting helps find the source of an impedance mismatch. Roughly
speaking, it encourages the rational programmer equip a module with types if it
is blamed in an error message. A sequence of such steps make up a portion of the
migratory path in the lattice of type migration~\cite{tfgnvf-popl-2016} (see
sec.~\ref{sub:stuff} for details).

@@

Hence the research problem is how to view the actions of a rational programmer
comparable across all semantics. Sections~\ref{sub:natural}
through~\ref{sub:erasure} describe the various rational programmers and explain
to what extent their actions are comparable.

@@

\rsub{stuff}     {Stuff}

\rsub{natural}   {The Natural Rational Programmer} 
\rsub{transient} {The Transient Rational Programmer} 
\rsub{erasure}   {The Erasure Rational Programmer}

>>>>

Section~\ref{sec:why-rational} explains, and illustrates with an example, how a
migratory type setting helps with finding the source of an impedance
mismatch. Roughly speaking, it encourages the rational programmer to equip a
module with types if it is blamed in an error message. A sequence of such steps
make up a portion of the migratory path in the lattice of type
migration~\cite{tfgnvf-popl-2016} (see sec.~\ref{sub:stuff} for details).

@@

Hence the research problem is how to make the actions of a rational programmer
comparable across all semantics. Sections~\ref{sub:natural}
through~\ref{sub:erasure} describe the various rational programmers and explain
to what extent their actions are comparable. Section\ref{sub:experiment}
enhances comparability with the programmer-effort dimension. Finally, the last
subsection states the experimental questions.

@@

\rsub{stuff}     {The Lattice and the Debugging Scenario} 
\rsub{natural}   {The Natural Rational Programmer} 
\rsub{transient} {The Transient Rational Programmer} 
\rsub{erasure}   {The Erasure Rational Programmer} 
\rsub{effort}    {The Programmer Effort} \label{subsec:effort}
\rsub{experiment}{The Experimental Questions}
